---
title: "Analysis"
format: html
---

```{r}
if(!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, gtsummary, kableExtra, broom, lme4, splines, ggeffects, nlme, splines, usmap, ggplot2, tidytext, usmap, gganimate, gifski, dplyr, lubridate, transformr, forcats, tidyr,zoo)
```


```{r}
load('data/dat.rda')
dat = dat |>
  mutate(date_num = as.numeric(date))
```

### Exploration Data Analysis
### Standardize region labels and colors
```{r}
dat |> 
  distinct(region, state) |> 
  group_by(region) |> 
  summarise(states = paste(sort(state), collapse = ", ")) 
```

```{r}
region_labels <- c(
  "New England", "Tri-State", "Mid-Atlantic", "Southeast",
  "Great Lakes", "South Central", "Plains",
  "Mountain", "West Coast", "Pacific NW"
)

dat <- dat |>
  mutate(region = factor(region, levels = 1:10, labels = region_labels))

region_colors_named <- c(
  "New England" = "#d95f02",
  "Tri-State" = "#bf812d",
  "Mid-Atlantic" = "#80b1d3",
  "Southeast" = "#5ab4ac",
  "Great Lakes" = "#1b9e77",
  "South Central" = "#66c2a5",
  "Plains" = "#3288bd",
  "Mountain" = "#984ea3",
  "West Coast" = "#e78ac3",
  "Pacific NW" = "#f781bf"
)
```

### show the regions and states on the US map
```{r}
plot_usmap(data = dat, values = "region", color = "white") +
  scale_fill_manual(values = region_colors_named, name = "Region") +
  theme_classic() +
  ggtitle("Regional Map of the US")

```

```{r}
## Animated US Map of Monthly COVID-19 Death Rates by State (2020–2023)

dat <- dat |> 
  mutate(death_rate = deaths / population * 100000)

dat_monthly <- dat |> 
  mutate(date_month = floor_date(date, "month")) |>
  group_by(state, date_month) |>
  summarise(death_rate = mean(death_rate, na.rm = TRUE), .groups = "drop")

p <- plot_usmap(data = dat_monthly, values = "death_rate", regions = "states") +
  scale_fill_gradient(low = "white", high = "red", name = "Deaths per 100K") +
  labs(
    title = "Monthly COVID-19 Death Rate by State",
    subtitle = "{format(frame_time, '%B %Y')}",
    caption = "Source: CDC",
    fill = "Death Rate"
  ) +
  theme_void(base_size = 13) +
  theme(legend.position = "right") +
  transition_time(date_month) +
  ease_aes("linear")

anim_obj <- animate(p, width = 1000, height = 700, fps = 5, renderer = gifski_renderer())

anim_save("us_death_rate.gif", animation = anim_obj)
```


###Describe if COVID-19 became less or more virulent across the different periods.(Analysis from pset4)

Plot a trend plot for cases, hospitalizations and deaths for each state. Color by region. Plot rates per $100,000$ people. Place the plots on top of each other. Hint: Use `pivot_longer` and `facet_wrap`.
```{r}
dat |> 
  pivot_longer(c(cases, deaths, hosp), names_to = "outcome",values_to = "rate") |> 
  mutate(rate = rate / pop * 100000) |>
  ggplot(aes(date, rate, group = state, color = region)) + geom_line(alpha = 0.25) + 
  facet_wrap(~outcome, ncol = 1, scales = 'free_y') + theme_classic()
```

### COVID-19 waves were automatically identified by combining smoothed case trajectories with relative week-over-week growth rates.
### Highlighted by Start-to-Next-Start Intervals
```{r}
dat_us <- dat |>
  group_by(date) |>
  summarise(cases = sum(cases, na.rm = TRUE), .groups = "drop") |>
  arrange(date) |>
  mutate(
    cases_smooth = rollmean(cases, k = 7, fill = NA, align = "center"),
    pct_change = (cases_smooth - lag(cases_smooth)) / lag(cases_smooth),
    high_growth = ifelse(pct_change > 0.10, 1, 0)
  )

label_waves <- function(x) {
  r <- rle(x)
  wave_id <- rep(NA, length(x))
  counter <- 1
  idx <- 1
  for (i in seq_along(r$lengths)) {
    if (r$values[i] == 1 && r$lengths[i] >= 2) {
      wave_id[idx:(idx + r$lengths[i] - 1)] <- counter
      counter <- counter + 1
    }
    idx <- idx + r$lengths[i]
  }
  return(wave_id)
}

dat_us <- dat_us |>
  mutate(
    wave = label_waves(high_growth)
  )

wave_starts <- dat_us |>
  filter(!is.na(wave)) |>
  group_by(wave) |>
  summarise(start_date = min(date), .groups = "drop") |>
  arrange(start_date)

wave_rects <- wave_starts |>
  mutate(
    label = format(start_date, "%b %d, %Y"),
    end_date = lead(start_date) - 1
  )

wave_rects$label <- factor(wave_rects$label, levels = wave_rects$label)

last_date <- max(dat_us$date, na.rm = TRUE)
wave_rects$end_date[nrow(wave_rects)] <- last_date

ggplot(dat_us, aes(x = date, y = cases_smooth)) +
  geom_rect(data = wave_rects,
            aes(xmin = start_date, xmax = end_date, ymin = -Inf, ymax = Inf, 
                fill = label),
            alpha = 0.15, inherit.aes = FALSE) +
  geom_line(color = "black") +
  geom_point(aes(color = as.factor(wave)), size = 1.5, alpha = 0.6) +
  scale_color_viridis_d(option = "D", na.translate = FALSE) +
  scale_fill_brewer(palette = "Set2", name = "Wave Start Date") +
  guides(color = "none") +
  labs(
    title = "COVID-19 Waves",
    x = "Date", y = "Smoothed Cases"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black")
  )
```
library(forcats)
library(ggplot2)
library(tidyverse)
library(viridis)
library(scales)

us_national_trend <- dat |> 
  group_by(date) |>
  summarise(
    pop = sum(pop, na.rm = TRUE),
    cases = sum(cases, na.rm = TRUE),
    deaths = sum(deaths, na.rm = TRUE),
    hosp = sum(hosp, na.rm = TRUE),
    vax = sum(series, na.rm = TRUE),
    booster = sum(booster, na.rm = TRUE)
  ) |> 
  mutate(
    case_rate = cases / pop * 100000,
    death_rate = deaths / pop * 100000,
    hosp_rate = hosp / pop * 100000,
    vax_rate = vax / pop * 100,
    booster_rate = booster / pop * 100
  ) |> 
  pivot_longer(
    cols = c(death_rate, vax_rate, booster_rate, case_rate, hosp_rate),
    names_to = "metric", values_to = "rate"
  ) |>
  mutate(metric = factor(metric, levels = c(
    "death_rate", "vax_rate", "booster_rate", "case_rate", "hosp_rate"
  )))

ggplot(us_national_trend, aes(x = date, y = rate, color = metric)) +
  geom_line(linewidth = 1) +
  scale_color_manual(
    values = c(
      "death_rate" = "#d62728",
      "vax_rate" = "goldenrod",
      "booster_rate" = "purple",
      "case_rate" = "#1f77b4",
      "hosp_rate" = "#2ca02c"
    ),
    labels = c(
      "Deaths per 100,000", "Vaccination Rate (%)", "Booster Rate (%)",
      "Cases per 100,000", "Hospitalizations per 100,000"
    )
  ) +
  facet_wrap(~ metric, ncol = 1, scales = "free_y", labeller = as_labeller(c(
    death_rate = "Deaths per 100,000",
    vax_rate = "Vaccination Rate (%)",
    booster_rate = "Booster Rate (%)",
    case_rate = "Cases per 100,000",
    hosp_rate = "Hospitalizations per 100,000"
  ))) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n%Y") +
  labs(
    title = "COVID-19 and Vaccination Trends in the United States (2020–2023)",
    x = "Date", y = "Rate",
    color = "",
    caption = "Source: CDC, US Census"
  ) +
  theme_classic(base_size = 13) +
  theme(axis.text.x = element_text(size = 9))

#us trend with vaccination and booster
dat$region_name <- factor(dat$region_name)

outcome_levels <- c("Deaths", "Vaccination Rate (%)", "Booster Rate (%)", "Cases", "Hospitalizations")

dat |>
  pivot_longer(c(cases, deaths, hosp, series, booster), names_to = "outcome", values_to = "value") |>
  mutate(
    rate = case_when(
      outcome %in% c("series", "booster") ~ value / pop * 100,
      TRUE ~ value / pop * 100000
    ),
    outcome = recode(outcome,
                     deaths = "Deaths",
                     series = "Vaccination Rate (%)",
                     booster = "Booster Rate (%)",
                     cases = "Cases",
                     hosp = "Hospitalizations"),
    outcome = factor(outcome, levels = outcome_levels)
  ) |>
  ggplot(aes(x = date, y = rate, group = state, color = region_name)) +
  geom_line(alpha = 0.7, linewidth = 0.5) +
  facet_wrap(~outcome, ncol = 1, scales = "free_y") +
  scale_color_viridis_d(option = "C", end = 0.9) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n%Y") +
  labs(
    title = "COVID-19 and Vaccination Trends by State and Region (2020–2023)",
    subtitle = "Rates per 100,000 population (vaccination/booster shown as %)",
    x = "Date", y = "Rate",
    color = "Region",
    caption = "Source: CDC, US Census"
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.x = element_text(size = 9),
    strip.text = element_text(face = "bold")
  )

```{r}
#COVID-19 Trends in the United States (2020–2023)
library(ggplot2)
library(tidyverse)
library(viridis)
library(scales)

us_national_trend <- dat |> 
  group_by(date) |>
  summarise(
    pop = sum(pop, na.rm = TRUE),
    cases = sum(cases, na.rm = TRUE),
    deaths = sum(deaths, na.rm = TRUE),
    hosp = sum(hosp, na.rm = TRUE)
  ) |> 
  pivot_longer(cols = c(cases, deaths, hosp), names_to = "outcome", values_to = "count") |>
  mutate(rate = count / pop * 100000)


ggplot(us_national_trend, aes(x = date, y = rate, color = outcome)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("cases" = "#1f77b4", "deaths" = "#d62728", "hosp" = "#2ca02c")) +
  facet_wrap(~ outcome, ncol = 1, scales = "free_y") +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n%Y") +
  labs(
    title = "COVID-19 Trends in the United States (2020–2023)",
    subtitle = "National rates per 100,000 population: cases, deaths, hospitalizations",
    x = "Date",
    y = "Rate per 100,000",
    color = "Outcome",
    caption = "Source: CDC, US Census"
  ) +
  theme_classic(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 9),
    legend.position = "none"
  )

#COVID-19 US Trends by State and Region (2020–2023)
dat$region_name <- factor(dat$region_name)
dat |> 
  pivot_longer(c(cases, deaths, hosp), names_to = "outcome", values_to = "rate") |> 
  mutate(rate = rate / pop * 100000) |> 
  ggplot(aes(x = date, y = rate, group = state, color = region_name)) +
  geom_line(alpha = 0.7, linewidth = 0.5) +  
  facet_wrap(~outcome, ncol = 1, scales = "free_y") +
  scale_color_viridis_d(option = "C", end = 0.9) + 
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n%Y") +
  labs(
    title = "COVID-19 US Trends by State and Region (2020–2023)",
    subtitle = "Rates per 100,000 population: cases, deaths, and hospitalizations",
    x = "Date",
    y = "Rate per 100,000",
    color = "Region",
    caption = "Source: CDC, US Census"
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 9),
    legend.position = "right"
  )
```

```{r}
dat_eda = dat |> 
  mutate(
    death_rate = deaths / pop * 1e5,
    case_rate = cases / pop * 1e5,
    hosp_rate = hosp / pop * 1e5,
    vax_rate = series / pop * 1e2,  # percent fully vaccinated
    booster_rate = booster / pop * 1e2  # percent boosted
  )

dat_eda = dat_eda |> 
  group_by(state) |> 
  arrange(date) |> 
  mutate(
    case_rate_lag2 = lag(case_rate, 2),
    hosp_rate_lag2 = lag(hosp_rate, 2)
  ) |> 
  ungroup()

dat_nofirst2_eda = dat_eda |>
  filter(!is.na(case_rate_lag2), !is.na(hosp_rate_lag2))

# 2D Histogram: Death Rate vs Predictors (No Lag)
dat_eda |> 
  select(death_rate, case_rate, hosp_rate, vax_rate, booster_rate) |>
  pivot_longer(cols = -death_rate, names_to = "predictor", values_to = "value") |>
  ggplot(aes(x = value, y = death_rate)) +
  geom_bin2d(bins = 80) +  
  facet_wrap(~ predictor, scales = "free_x") +
  scale_fill_viridis_c(option = "C") +
  labs(
    title = "2D Histogram: Death Rate vs Predictors (No Lag)",
    x = "Predictor Value",
    y = "Death Rate (per 100,000)"
  ) +
  theme_minimal()

# 2D Histogram: Death Rate vs Predictors (With Lagged Variables)
dat_nofirst2_eda |> 
  select(death_rate, case_rate_lag2, hosp_rate_lag2, vax_rate, booster_rate) |>
  pivot_longer(cols = -death_rate, names_to = "predictor", values_to = "value") |>
  ggplot(aes(x = value, y = death_rate)) +
  geom_bin2d(bins = 80) +  
  facet_wrap(~ predictor, scales = "free_x") +
  scale_fill_viridis_c(option = "C") +
  labs(
    title = "2D Histogram: Death Rate vs Predictors",
    x = "Predictor Value",
    y = "Death Rate (per 100,000)"
  ) +
  theme_minimal()

# Pairwise Plot: Death Rate vs Predictors (No Lag) using GGally::ggpairs
dat_nofirst2_eda |> 
  select(death_rate, case_rate, hosp_rate, vax_rate, booster_rate) |>
  GGally::ggpairs()

# Pairwise Plot: Death Rate vs Predictors (With Lagged Variables) using GGally::ggpairs
dat_nofirst2_eda |> 
  select(death_rate, case_rate_lag2, hosp_rate_lag2, vax_rate, booster_rate) |>
  GGally::ggpairs()
```

### Divide the pandemic period, 2020/01/20 to 2023/05/11(first confirmed case of COVID-19, to the end of the public health emergency) into waves. For each period compute the deaths rates by state. Describe which states did better or worse during the different periods.
```{r}
# Divide the pandemic period into waves (based on auto-detected wave_rects)
dat <- dat |> 
  mutate(wave = case_when(
    date >= as.Date("2020-02-22") & date <= as.Date("2020-06-05") ~ "2020/02/22 - 2020/06/05",
    date >= as.Date("2020-06-06") & date <= as.Date("2020-10-02") ~ "2020/06/06 - 2020/10/02",
    date >= as.Date("2020-10-03") & date <= as.Date("2021-06-18") ~ "2020/10/03 - 2021/06/18",
    date >= as.Date("2021-06-19") & date <= as.Date("2021-11-26") ~ "2021/06/19 - 2021/11/26",
    date >= as.Date("2021-11-27") & date <= as.Date("2022-04-08") ~ "2021/11/27 - 2022/04/08",
    date >= as.Date("2022-04-09") & date <= as.Date("2023-05-11") ~ "2022/04/09 - 2023/05/11",
    TRUE ~ NA_character_
  ))

death_summary <- dat |> 
  filter(!is.na(wave)) |> 
  group_by(state, wave, region) |> 
  summarise(total_deaths = sum(deaths, na.rm = TRUE),
            total_pop = mean(pop, na.rm = TRUE),
            death_rate = total_deaths / total_pop * 100000,
            .groups = "drop")

ggplot(death_summary, aes(x = reorder_within(state, death_rate, wave), 
                          y = death_rate, fill = region)) +
  geom_col() +
  facet_wrap(~wave, scales = "free_y") +
  coord_flip() +
  scale_x_reordered() + 
  labs(title = "COVID-19 Death Rate per 100,000 by State and Wave",
       x = "State", y = "Death Rate (per 100,000)") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 3))
```

To determine when vaccination started and when most of the population was vaccinated, compute the percent of the US population (including DC and Puerto Rico) vaccinated by date. Do the same for the booster. Then plot both percentages. (Analysis from pset4)
```{r}
p <- dat |> group_by(date) |> summarise(pop = sum(pop), series = sum(series),
                                   booster = sum(booster)) |>
  pivot_longer(c(booster, series), names_to = "vaccine", values_to = "rate") |>
  mutate(rate = rate / pop * 100) |>
  ggplot(aes(date, rate, color = vaccine)) + geom_line() + theme_classic() + 
  ylab('percent vaccinated') + ggtitle('COVID 19 US Vaccination rates over time')
print(p)
```

### Vaccination(series) vs death rate in wave2 (2020/07/01-2021/05/30)
```{r}
dat_wave2 <- dat |>
  filter(date >= as.Date("2020-07-01") & date <= as.Date("2021-05-30")) |>
  mutate(death_rate = deaths / pop * 100000) |>
  group_by(region, state) |>
  summarise(
    vax_rate = max(series, na.rm = TRUE) / max(pop, na.rm = TRUE) * 100,
    mean_death_rate = mean(death_rate, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(dat_wave2, aes(x = vax_rate, y = mean_death_rate, label = state, color = region)) +
  geom_point(alpha = 0.8, size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.8) +  
  ggrepel::geom_text_repel(size = 3) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Series Uptake vs. Average Death Rate in Wave 2 (2020-07 to 2021-05)",
    subtitle = "Alpha variant period — boosters began rolling out",
    x = "Vaccination Rate by End of Wave (%)",
    y = "Average Death Rate (per 100k)",
    color = "Region"
  )
```
### Vaccination(booster) vs death rate in wave3 (2021/06/01-2021/10/31)

```{r}
dat_wave3 <- dat |>
  filter(date >= as.Date("2021-06-01") & date <= as.Date("2021-10-31")) |>
   mutate(death_rate = deaths / pop * 100000) |>
  group_by(region, state) |>
  summarise(
    booster_rate = max(booster, na.rm = TRUE) / max(pop, na.rm = TRUE) * 100,
    mean_death_rate = mean(death_rate, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(dat_wave3, aes(x = booster_rate, y = mean_death_rate, label = state, color = region)) +
  geom_point(alpha = 0.8, size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.8) +
  ggrepel::geom_text_repel(size = 3) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Booster Uptake vs. Average Death Rate in Wave 3 (2021-06 to 2021-10)",
    subtitle = "Delta variant period — boosters began rolling out",
    x = "Booster Rate by End of Wave (%)",
    y = "Average Death Rate (per 100k)",
    color = "Region"
  )
```

```{r}
#vaccination
dat_eda = dat |> 
  mutate(
    death_rate = deaths / pop * 1e5,
    case_rate = cases / pop * 1e5,
    hosp_rate = hosp / pop * 1e5,
    vax_rate = series / pop * 1e2,  # percent fully vaccinated
    booster_rate = booster / pop * 1e2  # percent boosted
  )

dat_eda = dat_eda |> 
  group_by(state) |> 
  arrange(date) |> 
  mutate(
    case_rate_lag2 = lag(case_rate, 2),
    hosp_rate_lag2 = lag(hosp_rate, 2)
  ) |> 
  ungroup()

dat_nofirst2_eda = dat_eda |>
  filter(!is.na(case_rate_lag2), !is.na(hosp_rate_lag2))

# Heatmap of Vaccination Rate by State and Month (States Sorted by Average Vaccination Rate)
state_vax_avg <- dat_nofirst2_eda |>
  group_by(state) |>
  summarise(vax_avg = mean(vax_rate, na.rm = TRUE))

dat_nofirst2_eda |>
  mutate(month = floor_date(date, "month")) |>
  left_join(state_vax_avg, by = "state") |>  
  mutate(state = fct_reorder(state, vax_avg)) |>  
  group_by(state, month) |>
  summarise(
    death_rate = mean(death_rate, na.rm = TRUE),
    vax_rate = mean(vax_rate, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = month, y = state, fill = vax_rate)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(name = "Vax Rate") +
  labs(
    title = "Vaccination Rate Heatmap by State and Month",
    x = "Month", y = "State (sorted by avg. Vax Rate)"
  ) +
  theme_minimal()

# Heatmap of Booster Rate by State and Month (States Sorted by Average Booster Rate)
state_booster_avg <- dat_nofirst2_eda |>
  group_by(state) |>
  summarise(booster_avg = mean(booster_rate, na.rm = TRUE))

dat_nofirst2_eda |>
  mutate(month = floor_date(date, "month")) |>
  left_join(state_booster_avg, by = "state") |>
  mutate(state = fct_reorder(state, booster_avg)) |>
  group_by(state, month) |>
  summarise(
    booster_rate = mean(booster_rate, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = month, y = state, fill = booster_rate)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(name = "Booster Rate") +
  labs(
    title = "Booster Vaccination Rate Heatmap by State and Month",
    x = "Month", y = "State (sorted by avg. Booster Rate)"
  ) +
  theme_minimal()

# Heatmap of Vaccination Rate by State, Month, and Region (Faceted by Region)
state_vax_avg <- dat_nofirst2_eda |>
  group_by(state) |>
  summarise(vax_avg = mean(vax_rate, na.rm = TRUE))

dat_nofirst2_eda |>
  mutate(month = floor_date(date, "month")) |>
  left_join(state_vax_avg, by = "state") |>
  mutate(state = fct_reorder(state, vax_avg)) |>
  group_by(state, month, region_name) |>
  summarise(
    vax_rate = mean(vax_rate, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = month, y = state, fill = vax_rate)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(name = "Vax Rate") +
  labs(
    title = "Vaccination Rate Heatmap by State and Month",
    x = "Month", y = "State"
  ) +
  facet_wrap(~ region_name, scales = "free_y") +
  theme_minimal(base_size = 12)

# Scatter Plot of Death Rate vs Vaccination and Booster Rates (with LOESS Smoothing)
dat_nofirst2_eda |> 
  select(date, state, death_rate, vax_rate, booster_rate) |> 
  pivot_longer(cols = c(vax_rate, booster_rate), names_to = "type", values_to = "rate") |>
  ggplot(aes(x = rate, y = death_rate)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  facet_wrap(~type, scales = "free_x") +
  labs(
    title = "Relationship Between Vaccination Rates and Death Rate",
    x = "Vaccination / Booster Rate (%)",
    y = "Death Rate (per 100,000)"
  ) +
  theme_minimal()

# Bar Plot of Average Death Rate by Vaccination Rate Group
dat_nofirst2_eda |>
  mutate(
    vax_group = cut(vax_rate, breaks = c(0, 25, 50, 75, 100), include.lowest = TRUE),
    booster_group = cut(booster_rate, breaks = c(0, 25, 50, 75, 100), include.lowest = TRUE)
  ) |>
  group_by(vax_group) |>
  summarise(mean_death = mean(death_rate, na.rm = TRUE)) |>
  ggplot(aes(x = vax_group, y = mean_death)) +
  geom_col(fill = "blue") +
  labs(
    title = "Average Death Rate by Vaccination Group",
    x = "Vaccination Rate Group (%)",
    y = "Mean Death Rate (per 100,000)"
  ) +
  theme_minimal()

```

### LME Models

#### Prepare variables
```{r}
dat = dat |> 
  mutate(
    death_rate = deaths / pop * 1e5,
    case_rate = cases / pop * 1e5,
    hosp_rate = hosp / pop * 1e5,
    vax_rate = series / pop * 1e2,  # percent fully vaccinated
    booster_rate = booster / pop * 1e2  # percent boosted
  )

dat = dat |> 
  group_by(state) |> 
  arrange(date) |> 
  mutate(
    case_rate_lag2 = lag(case_rate, 2),
    hosp_rate_lag2 = lag(hosp_rate, 2)
  ) |> 
  ungroup()

dat_nofirst2 = dat |>
  filter(!is.na(case_rate_lag2), !is.na(hosp_rate_lag2))
```


##### Model 1 lmer
```{r}
model_1 = lmer(death_rate ~ case_rate + hosp_rate + vax_rate + booster_rate + ns(date_num, df = 5) + (1 | state), data = dat_nofirst2)

summary(model_1)


plot(ggpredict(model_1, terms="date_num [all]"))
AIC(model_1)
```



##### Model 2 lag cases hosp
```{r}
model_lag = lmer(death_rate ~ case_rate_lag2 + hosp_rate_lag2 + vax_rate + booster_rate + ns(date_num, df = 8) + (1 | state), data = dat_nofirst2)

summary(model_lag)

plot(ggpredict(model_lag, terms="date_num [all]"))
AIC(model_lag)
```

```{r}
aic_results = tibble(df = 2:10, AIC = NA_real_)

# Loop over degrees of freedom
for (i in 2:10) {
  model = lmer(death_rate ~ case_rate_lag2 + hosp_rate_lag2 + vax_rate + booster_rate +
                 ns(date_num, df = i) + (1 | state), data = dat_nofirst2)
  aic_results$AIC[aic_results$df == i] = AIC(model)
}

print(aic_results)
which.min(aic_results$AIC)
```



```{r}
fit = model_lag
state_to_plot = "CA"

# Remove rows with missing lagged predictors before prediction
dat_pred = dat |>
  filter(!is.na(case_rate_lag2), !is.na(hosp_rate_lag2)) |>
  mutate(predicted = predict(fit))

# Join predictions back to full data (optional, only if you want to keep NA rows in plotting)
dat_state = dat |> 
  filter(state == state_to_plot) |>
  left_join(dat_pred |> select(state, date, predicted), by = c("state", "date"))

# Plot observed vs predicted
ggplot(dat_state, aes(x = date)) +
  geom_point(aes(y = death_rate), color = "red", alpha = 0.6, size = 1.2) +  # Observed
  geom_line(aes(y = predicted), color = "blue", linewidth = 1) +  # Predicted
  labs(title = paste("Observed vs. Predicted Death Rate in", state_to_plot),
       y = "Death Rate (per 100,000)",
       x = "Date") +
  theme_minimal()
```

### Model Diagnostics
```{r}
plot(model_lag)  # base R method for lmer

resid_df = data.frame(
  fitted = fitted(model_lag),
  residuals = resid(model_lag)
)

ggplot(resid_df, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  theme_minimal() +
  labs(title = "Residuals vs. Fitted", x = "Fitted", y = "Residuals")
```

```{r}
qqnorm(resid(model_lag))
qqline(resid(model_lag), col = "blue")

data.frame(resid = resid(model_lag)) |> 
  ggplot(aes(sample = resid)) +
  stat_qq() +
  stat_qq_line(color = "blue") +
  theme_minimal() +
  labs(title = "Q-Q Plot of Residuals")
```data:image/png;base64,iVBORw0KGgoAAAANSUhEUgAAAAIAAAAUCAYAAACnOeyiAAAAXklEQVR4XmNgAIL///8zMYSGhjIDGYIMIiIMvECGMwMDN4M4kFEDUqIIZKwDMdSBjAsghj6Q8QPEMAAy/lOBoQekv4AYKkDGfgZeXl4RICOLQUtLiw3IUAJJMQIZ7AC2tU2tJWCy/wAAAABJRU5ErkJggg==

### Excess death rate
```{r}
# Predict death rate using the lag model
dat_excess = dat_nofirst2 |> 
  mutate(predicted_death_rate = predict(model_lag),
         excess_death_rate = death_rate - predicted_death_rate)

# View a few rows
head(dat_excess |> select(state, date, death_rate, predicted_death
```_rate, excess_death_rate))

```{r}
state_to_plot = "CA"

dat_excess |> 
  filter(state == state_to_plot) |> 
  ggplot(aes(x = date, y = excess_death_rate)) +
  geom_line(color = "darkred") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_minimal() +
  labs(title = paste("Excess Death Rate Over Time in", state_to_plot),
       y = "Excess Death Rate (per 100,000)",
       x = "Date")
```


