---
title: "Analysis"
format: html
---

```{r}
if(!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, gtsummary, kableExtra, broom, lme4, splines, ggeffects, nlme, usmap, ggplot2, tidytext, gganimate, gifski, dplyr, lubridate, transformr, forcats, tidyr, zoo, colorspace, purrr)
```


```{r}
load('data/dat.rda')
dat = dat |>
  mutate(date_num = as.numeric(date))
```

### Exploration Data Analysis
### Standardize region labels and colors
```{r}
dat |> 
  distinct(region, state) |> 
  group_by(region) |> 
  summarise(states = paste(sort(state), collapse = ", ")) 
```

```{r}
region_labels <- c(
  "New England", "Tri-State", "Mid-Atlantic", "Southeast",
  "Great Lakes", "South Central", "Plains",
  "Mountain", "West Coast", "Pacific NW"
)

dat <- dat |>
  mutate(region = factor(region, levels = 1:10, labels = region_labels))

region_colors_named <- c(
  "New England" = "#d95f02",
  "Tri-State" = "#bf812d",
  "Mid-Atlantic" = "#80b1d3",
  "Southeast" = "#5ab4ac",
  "Great Lakes" = "#1b9e77",
  "South Central" = "#66c2a5",
  "Plains" = "#3288bd",
  "Mountain" = "#984ea3",
  "West Coast" = "#e78ac3",
  "Pacific NW" = "#f781bf"
)
```

### show the regions and states on the US map
```{r USmap}
plot_usmap(data = dat, values = "region", color = "white") +
  scale_fill_manual(values = region_colors_named, name = "Region") +
  theme_classic() +
  ggtitle("Regional Map of the US")

```

```{r gif_animation, eval=FALSE}
## Animated US Map of Monthly COVID-19 Death Rates by State (2020–2023)

dat <- dat |> 
  mutate(death_rate = deaths / population * 100000)

dat_monthly <- dat |> 
  mutate(date_month = floor_date(date, "month")) |>
  group_by(state, date_month) |>
  summarise(death_rate = mean(death_rate, na.rm = TRUE), .groups = "drop")

p <- plot_usmap(data = dat_monthly, values = "death_rate", regions = "states") +
  scale_fill_gradient(low = "white", high = "red", name = "Deaths per 100K") +
  labs(
    title = "Monthly COVID-19 Death Rate by State",
    subtitle = "{format(frame_time, '%B %Y')}",
    caption = "Source: CDC",
    fill = "Death Rate"
  ) +
  theme_void(base_size = 13) +
  theme(legend.position = "right") +
  transition_time(date_month) +
  ease_aes("linear")

anim_obj <- animate(p, width = 1000, height = 700, fps = 5, renderer = gifski_renderer())

anim_save("us_death_rate.gif", animation = anim_obj)
```


###Describe if COVID-19 became less or more virulent across the different periods.(Analysis from pset4)

Plot a trend plot for cases, hospitalizations and deaths for each state. Color by region. Plot rates per $100,000$ people. Place the plots on top of each other. Hint: Use `pivot_longer` and `facet_wrap`.
```{r trend_plot_for_cases_hosp_deaths}
dat |> 
  pivot_longer(c(cases, deaths, hosp), names_to = "outcome",values_to = "rate") |> 
  mutate(rate = rate / pop * 100000) |>
  ggplot(aes(date, rate, group = state, color = region)) + geom_line(alpha = 0.25) + 
  facet_wrap(~outcome, ncol = 1, scales = 'free_y') + theme_classic()
```

### COVID-19 waves were automatically identified by combining smoothed case trajectories with relative week-over-week growth rates.
### Highlighted by Start-to-Next-Start Intervals
```{r death_wave_Identification, fig.width = 16, fig.height = 6, message=FALSE, warning=FALSE}
dat_us <- dat |>
  group_by(date) |>
  summarise(deaths = sum(deaths, na.rm = TRUE), .groups = "drop") |>
  arrange(date) |>
  mutate(
    deaths_smooth = rollmean(deaths, k = 7, fill = NA, align = "center"),
    pct_change = (deaths_smooth - lag(deaths_smooth)) / lag(deaths_smooth),
    high_growth = ifelse(pct_change > 0.10, 1, 0)
  )

label_waves <- function(x) {
  r <- rle(x)
  wave_id <- rep(NA, length(x))
  counter <- 1
  idx <- 1
  for (i in seq_along(r$lengths)) {
    if (r$values[i] == 1 && r$lengths[i] >= 2) {
      wave_id[idx:(idx + r$lengths[i] - 1)] <- counter
      counter <- counter + 1
    }
    idx <- idx + r$lengths[i]
  }
  return(wave_id)
}

dat_us <- dat_us |>
  mutate(
    wave = label_waves(high_growth)
  )

wave_starts <- dat_us |>
  filter(!is.na(wave)) |>
  group_by(wave) |>
  summarise(start_date = min(date), .groups = "drop") |>
  arrange(start_date)

wave_rects <- wave_starts |>
  mutate(
    label = format(start_date, "%b %d, %Y"),
    end_date = lead(start_date) - 1
  )

wave_rects$label <- factor(wave_rects$label, levels = wave_rects$label)

last_date <- max(dat_us$date, na.rm = TRUE)
wave_rects$end_date[nrow(wave_rects)] <- last_date

ggplot(dat_us, aes(x = date, y = deaths_smooth)) +
  geom_rect(data = wave_rects,
            aes(xmin = start_date, xmax = end_date, ymin = -Inf, ymax = Inf, 
                fill = label),
            alpha = 0.15, inherit.aes = FALSE) +
  geom_line(color = "black") +
  geom_point(aes(color = as.factor(wave)), size = 1.5, alpha = 0.6) +
  scale_color_viridis_d(option = "D", na.translate = FALSE) +
  scale_fill_brewer(name = "Wave Start Date") +
  guides(color = "none") +
  labs(
    title = "COVID-19 Waves",
    x = "Date", y = "Smoothed deaths"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black")
  )
```
```{r wave_plot_with_standardized_trends, fig.width = 16, fig.height = 6, message=FALSE, warning=FALSE}
dat_us <- dat |>
  group_by(date) |>
  summarise(
    cases = sum(cases, na.rm = TRUE),
    hosp = sum(hosp, na.rm = TRUE),
    deaths = sum(deaths, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(date) |>
  mutate(
    cases_smooth = rollmean(cases, k = 7, fill = NA, align = "center"),
    hosp_smooth = rollmean(hosp, k = 7, fill = NA, align = "center"),
    deaths_smooth = rollmean(deaths, k = 7, fill = NA, align = "center"),

    cases_std = cases_smooth / max(cases_smooth, na.rm = TRUE),
    hosp_std = hosp_smooth / max(hosp_smooth, na.rm = TRUE),
    deaths_std = deaths_smooth / max(deaths_smooth, na.rm = TRUE),

    pct_change = (deaths_smooth - lag(deaths_smooth)) / lag(deaths_smooth),
    high_growth = ifelse(pct_change > 0.10, 1, 0)
  )

label_waves <- function(x) {
  r <- rle(x)
  wave_id <- rep(NA, length(x))
  counter <- 1
  idx <- 1
  for (i in seq_along(r$lengths)) {
    if (r$values[i] == 1 && r$lengths[i] >= 2) {
      wave_id[idx:(idx + r$lengths[i] - 1)] <- counter
      counter <- counter + 1
    }
    idx <- idx + r$lengths[i]
  }
  return(wave_id)
}

dat_us <- dat_us |>
  mutate(wave = label_waves(high_growth))

wave_starts <- dat_us |>
  filter(!is.na(wave)) |>
  group_by(wave) |>
  summarise(start_date = min(date), .groups = "drop") |>
  arrange(start_date)

wave_rects <- wave_starts |>
  mutate(
    label = format(start_date, "%b %d, %Y"),
    end_date = lead(start_date) - 1
  )

last_date <- max(dat_us$date, na.rm = TRUE)
wave_rects$end_date[nrow(wave_rects)] <- last_date
wave_rects$label <- factor(wave_rects$label, levels = wave_rects$label)

ggplot(dat_us, aes(x = date)) +
  geom_rect(data = wave_rects,
            aes(xmin = start_date, xmax = end_date, ymin = -Inf, ymax = Inf, fill = label),
            alpha = 0.15, inherit.aes = FALSE) +
  geom_line(aes(y = cases_std, color = "Cases"), linewidth = 0.8) +
  geom_line(aes(y = hosp_std, color = "Hospitalizations"), linewidth = 0.8) +
  geom_line(aes(y = deaths_std, color = "Deaths"), linewidth = 1.2) +
  scale_color_manual(values = c("Cases" = "blue", "Hospitalizations" = "darkgreen", "Deaths" = "black")) +
  scale_fill_brewer(name = "Wave Start Date") +
  labs(
    title = "COVID-19 Waves with Standardized Smoothed Trends",
    x = "Date",
    y = "Relative Level (Standardized 0–1)",
    color = "Metric"
  ) +
  theme_minimal(base_size = 14) +
  guides(fill = guide_legend(order = 1), color = guide_legend(order = 2)) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black")
  )

vax_df <- dat |>
  group_by(date) |>
  summarise(pop = sum(pop), 
            series = sum(series, na.rm = TRUE),
            booster = sum(booster, na.rm = TRUE),
            .groups = "drop") |>
  mutate(series_pct = series / pop * 100,
         booster_pct = booster / pop * 100)

# 合并疫苗数据到 dat_us
dat_us <- left_join(dat_us, vax_df |> select(date, series_pct, booster_pct), by = "date")

# 绘图
ggplot(dat_us, aes(x = date)) +
  # 波段背景
  geom_rect(data = wave_rects,
            aes(xmin = start_date, xmax = end_date, ymin = -Inf, ymax = Inf, fill = label),
            alpha = 0.15, inherit.aes = FALSE) +
  
  # 标准化曲线
  geom_line(aes(y = cases_std, color = "Cases"), linewidth = 0.8) +
  geom_line(aes(y = hosp_std, color = "Hospitalizations"), linewidth = 0.8) +
  geom_line(aes(y = deaths_std, color = "Deaths"), linewidth = 1.2) +

  # 疫苗接种趋势（用第二y轴比例缩放）
  geom_line(aes(y = series_pct / 100, color = "Primary Series"), linewidth = 1, linetype = "dashed") +
  geom_line(aes(y = booster_pct / 100, color = "Booster"), linewidth = 1, linetype = "dotted") +

  # 设置颜色
  scale_color_manual(
    name = "Metric",
    values = c("Cases" = "blue", 
               "Hospitalizations" = "darkgreen", 
               "Deaths" = "black",
               "Primary Series" = "orange",
               "Booster" = "purple")
  ) +

  # 填充图例
  scale_fill_brewer(name = "Wave Start Date") +

  # 坐标轴与标题
  labs(
    title = "COVID-19 Waves with Standardized Trends and Vaccination Rates",
    x = "Date",
    y = "Relative Level (Standardized 0–1)",
    caption = "Vaccination rates shown as % / 100 for overlaying"
  ) +

  # 图形样式
  theme_minimal(base_size = 14) +
  guides(fill = guide_legend(order = 1), color = guide_legend(order = 2)) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black")
  )
```

```{r}
#COVID-19 Trends in the United States (2020–2023)
library(ggplot2)
library(tidyverse)
library(viridis)
library(scales)

us_national_trend <- dat |> 
  group_by(date) |>
  summarise(
    pop = sum(pop, na.rm = TRUE),
    cases = sum(cases, na.rm = TRUE),
    deaths = sum(deaths, na.rm = TRUE),
    hosp = sum(hosp, na.rm = TRUE)
  ) |> 
  pivot_longer(cols = c(cases, deaths, hosp), names_to = "outcome", values_to = "count") |>
  mutate(rate = count / pop * 100000)


ggplot(us_national_trend, aes(x = date, y = rate, color = outcome)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("cases" = "#1f77b4", "deaths" = "#d62728", "hosp" = "#2ca02c")) +
  facet_wrap(~ outcome, ncol = 1, scales = "free_y") +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n%Y") +
  labs(
    title = "COVID-19 Trends in the United States (2020–2023)",
    subtitle = "National rates per 100,000 population: cases, deaths, hospitalizations",
    x = "Date",
    y = "Rate per 100,000",
    color = "Outcome",
    caption = "Source: CDC, US Census"
  ) +
  theme_classic(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 9),
    legend.position = "none"
  )

#COVID-19 US Trends by State and Region (2020–2023)
dat$region_name <- factor(dat$region_name)
dat |> 
  pivot_longer(c(cases, deaths, hosp), names_to = "outcome", values_to = "rate") |> 
  mutate(rate = rate / pop * 100000) |> 
  ggplot(aes(x = date, y = rate, group = state, color = region_name)) +
  geom_line(alpha = 0.7, linewidth = 0.5) +  
  facet_wrap(~outcome, ncol = 1, scales = "free_y") +
  scale_color_viridis_d(option = "C", end = 0.9) + 
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n%Y") +
  labs(
    title = "COVID-19 US Trends by State and Region (2020–2023)",
    subtitle = "Rates per 100,000 population: cases, deaths, and hospitalizations",
    x = "Date",
    y = "Rate per 100,000",
    color = "Region",
    caption = "Source: CDC, US Census"
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 9),
    legend.position = "right"
  )
```

```{r}
dat_eda = dat |> 
  mutate(
    death_rate = deaths / pop * 1e5,
    case_rate = cases / pop * 1e5,
    hosp_rate = hosp / pop * 1e5,
    vax_rate = series / pop * 1e2,  # percent fully vaccinated
    booster_rate = booster / pop * 1e2  # percent boosted
  )

dat_eda = dat_eda |> 
  group_by(state) |> 
  arrange(date) |> 
  mutate(
    case_rate_lag2 = lag(case_rate, 2),
    hosp_rate_lag2 = lag(hosp_rate, 2)
  ) |> 
  ungroup()

dat_nofirst2_eda = dat_eda |>
  filter(!is.na(case_rate_lag2), !is.na(hosp_rate_lag2))

# 2D Histogram: Death Rate vs Predictors (No Lag)
dat_eda |> 
  select(death_rate, case_rate, hosp_rate, vax_rate, booster_rate) |>
  pivot_longer(cols = -death_rate, names_to = "predictor", values_to = "value") |>
  ggplot(aes(x = value, y = death_rate)) +
  geom_bin2d(bins = 80) +  
  facet_wrap(~ predictor, scales = "free_x") +
  scale_fill_viridis_c(option = "C") +
  labs(
    title = "2D Histogram: Death Rate vs Predictors (No Lag)",
    x = "Predictor Value",
    y = "Death Rate (per 100,000)"
  ) +
  theme_minimal()

# 2D Histogram: Death Rate vs Predictors (With Lagged Variables)
dat_nofirst2_eda |> 
  select(death_rate, case_rate_lag2, hosp_rate_lag2, vax_rate, booster_rate) |>
  pivot_longer(cols = -death_rate, names_to = "predictor", values_to = "value") |>
  ggplot(aes(x = value, y = death_rate)) +
  geom_bin2d(bins = 80) +  
  facet_wrap(~ predictor, scales = "free_x") +
  scale_fill_viridis_c(option = "C") +
  labs(
    title = "2D Histogram: Death Rate vs Predictors",
    x = "Predictor Value",
    y = "Death Rate (per 100,000)"
  ) +
  theme_minimal()

# Pairwise Plot: Death Rate vs Predictors (No Lag) using GGally::ggpairs
dat_nofirst2_eda |> 
  select(death_rate, case_rate, hosp_rate, vax_rate, booster_rate) |>
  GGally::ggpairs()

# Pairwise Plot: Death Rate vs Predictors (With Lagged Variables) using GGally::ggpairs
dat_nofirst2_eda |> 
  select(death_rate, case_rate_lag2, hosp_rate_lag2, vax_rate, booster_rate) |>
  GGally::ggpairs()
```

### Divide the pandemic period, 2020/01/20 to 2023/05/11(first confirmed case of COVID-19, to the end of the public health emergency) into waves. For each period compute the deaths rates by state. Describe which states did better or worse during the different periods.
```{r wave_death_rate, fig.width = 16, fig.height = 6, message=FALSE, warning=FALSE}

wave_rects_labeled <- wave_rects |>
  select(wave_label = label, start_date, end_date)


dat <- dat |> 
  left_join(wave_rects_labeled, by = character()) |> 
  filter(date >= start_date & date <= end_date) |> 
  mutate(wave = wave_label) |> 
  select(-start_date, -end_date, -wave_label)

death_summary <- dat |> 
  filter(!is.na(wave)) |> 
  group_by(state, wave, region) |> 
  summarise(total_deaths = sum(deaths, na.rm = TRUE),
            total_pop = mean(pop, na.rm = TRUE),
            death_rate = total_deaths / total_pop * 100000,
            .groups = "drop")

ggplot(death_summary, aes(x = reorder_within(state, death_rate, wave), 
                          y = death_rate, fill = region)) +
  geom_col() +
  facet_wrap(~wave, scales = "free_y", nrow = 1) +
  coord_flip() +
  scale_x_reordered() + 
  labs(title = "COVID-19 Death Rate by State and Wave",
       x = "State", y = "Death Rate (per 100,000)") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 3))
```

To determine when vaccination started and when most of the population was vaccinated, compute the percent of the US population (including DC and Puerto Rico) vaccinated by date. Do the same for the booster. Then plot both percentages. (Analysis from pset4)
```{r}
p <- dat |> group_by(date) |> summarise(pop = sum(pop), series = sum(series),
                                   booster = sum(booster)) |>
  pivot_longer(c(booster, series), names_to = "vaccine", values_to = "rate") |>
  mutate(rate = rate / pop * 100) |>
  ggplot(aes(date, rate, color = vaccine)) + geom_line() + theme_classic() + 
  ylab('percent vaccinated') + ggtitle('COVID 19 US Vaccination rates over time')
print(p)
```

### Vaccination(series) vs death rate in wave4
```{r}
wave4_dates <- wave_rects[4, ]

dat_wave4 <- dat |>
  filter(date >= wave4_dates$start_date & date <= wave4_dates$end_date) |>
  mutate(death_rate = deaths / pop * 100000) |>
  group_by(region, state) |>
  summarise(
    vax_rate = max(series, na.rm = TRUE) / max(pop, na.rm = TRUE) * 100,
    mean_death_rate = mean(death_rate, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(dat_wave4, aes(x = vax_rate, y = mean_death_rate, label = state, color = region)) +
  geom_point(alpha = 0.8, size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.8) +  
  ggrepel::geom_text_repel(size = 3) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Series Uptake vs. Average Death Rate in Wave 4",
    x = "Vaccination Rate by End of Wave (%)",
    y = "Average Death Rate (per 100k)",
    color = "Region"
  )
```
### Vaccination(booster) vs death rate in wave3 (2021/06/01-2021/10/31)

```{r}
wave4_dates <- wave_rects[4, ]

dat_wave4 <- dat |>
  filter(date >= wave3_dates$start_date & date <= wave3_dates$end_date) |>
  mutate(death_rate = deaths / pop * 100000) |>
  group_by(region, state) |>
  summarise(
    vax_rate = max(series, na.rm = TRUE) / max(pop, na.rm = TRUE) * 100,
    mean_death_rate = mean(death_rate, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(dat_wave4, aes(x = vax_rate, y = mean_death_rate, label = state, color = region)) +
  geom_point(alpha = 0.8, size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.8) +  
  ggrepel::geom_text_repel(size = 3) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Booster Uptake vs. Average Death Rate in Wave 4",
    x = "Vaccination Rate by End of Wave (%)",
    y = "Average Death Rate (per 100k)",
    color = "Region"
  )
```

```{r}
#vaccination
dat_eda = dat |> 
  mutate(
    death_rate = deaths / pop * 1e5,
    case_rate = cases / pop * 1e5,
    hosp_rate = hosp / pop * 1e5,
    vax_rate = series / pop * 1e2,  # percent fully vaccinated
    booster_rate = booster / pop * 1e2  # percent boosted
  )

dat_eda = dat_eda |> 
  group_by(state) |> 
  arrange(date) |> 
  mutate(
    case_rate_lag2 = lag(case_rate, 2),
    hosp_rate_lag2 = lag(hosp_rate, 2)
  ) |> 
  ungroup()

dat_nofirst2_eda = dat_eda |>
  filter(!is.na(case_rate_lag2), !is.na(hosp_rate_lag2))

# Heatmap of Vaccination Rate by State and Month (States Sorted by Average Vaccination Rate)
state_vax_avg <- dat_nofirst2_eda |>
  group_by(state) |>
  summarise(vax_avg = mean(vax_rate, na.rm = TRUE))

dat_nofirst2_eda |>
  mutate(month = floor_date(date, "month")) |>
  left_join(state_vax_avg, by = "state") |>  
  mutate(state = fct_reorder(state, vax_avg)) |>  
  group_by(state, month) |>
  summarise(
    death_rate = mean(death_rate, na.rm = TRUE),
    vax_rate = mean(vax_rate, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = month, y = state, fill = vax_rate)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(name = "Vax Rate") +
  labs(
    title = "Vaccination Rate Heatmap by State and Month",
    x = "Month", y = "State (sorted by avg. Vax Rate)"
  ) +
  theme_minimal()

# Heatmap of Booster Rate by State and Month (States Sorted by Average Booster Rate)
state_booster_avg <- dat_nofirst2_eda |>
  group_by(state) |>
  summarise(booster_avg = mean(booster_rate, na.rm = TRUE))

dat_nofirst2_eda |>
  mutate(month = floor_date(date, "month")) |>
  left_join(state_booster_avg, by = "state") |>
  mutate(state = fct_reorder(state, booster_avg)) |>
  group_by(state, month) |>
  summarise(
    booster_rate = mean(booster_rate, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = month, y = state, fill = booster_rate)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(name = "Booster Rate") +
  labs(
    title = "Booster Vaccination Rate Heatmap by State and Month",
    x = "Month", y = "State (sorted by avg. Booster Rate)"
  ) +
  theme_minimal()



# Heatmap of Vaccination Rate by State, Month, and Region (Faceted by Region)
state_vax_avg <- dat_nofirst2_eda |>
  group_by(state) |>
  summarise(vax_avg = mean(vax_rate, na.rm = TRUE))

dat_nofirst2_eda |>
  mutate(month = floor_date(date, "month")) |>
  left_join(state_vax_avg, by = "state") |>
  mutate(state = fct_reorder(state, vax_avg)) |>
  group_by(state, month, region_name) |>
  summarise(
    vax_rate = mean(vax_rate, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = month, y = state, fill = vax_rate)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(name = "Vax Rate") +
  labs(
    title = "Vaccination Rate Heatmap by State and Month",
    x = "Month", y = "State"
  ) +
  facet_wrap(~ region_name, scales = "free_y") +
  theme_minimal(base_size = 12)

# Scatter Plot of Death Rate vs Vaccination and Booster Rates (with LOESS Smoothing)
dat_nofirst2_eda |> 
  select(date, state, death_rate, vax_rate, booster_rate) |> 
  pivot_longer(cols = c(vax_rate, booster_rate), names_to = "type", values_to = "rate") |>
  ggplot(aes(x = rate, y = death_rate)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  facet_wrap(~type, scales = "free_x") +
  labs(
    title = "Relationship Between Vaccination Rates and Death Rate",
    x = "Vaccination / Booster Rate (%)",
    y = "Death Rate (per 100,000)"
  ) +
  theme_minimal()

# Bar Plot of Average Death Rate by Vaccination Rate Group
dat_nofirst2_eda |>
  mutate(
    vax_group = cut(vax_rate, breaks = c(0, 25, 50, 75, 100), include.lowest = TRUE),
    booster_group = cut(booster_rate, breaks = c(0, 25, 50, 75, 100), include.lowest = TRUE)
  ) |>
  group_by(vax_group) |>
  summarise(mean_death = mean(death_rate, na.rm = TRUE)) |>
  ggplot(aes(x = vax_group, y = mean_death)) +
  geom_col(fill = "blue") +
  labs(
    title = "Average Death Rate by Vaccination Group",
    x = "Vaccination Rate Group (%)",
    y = "Mean Death Rate (per 100,000)"
  ) +
  theme_minimal()

```
```{r}
region_palettes <- map(region_colors_named, ~ lighten(.x, seq(0.2, 1, length.out = 9)))
booster_plot_data <- dat_nofirst2_eda |>
  mutate(
    month = floor_date(date, "month")
  ) |>
  left_join(state_booster_avg, by = "state") |>
  left_join(select(dat, state, region), by = "state") |> 
  mutate(state = fct_reorder(state, booster_avg)) |>
  group_by(state, region, month) |>
  summarise(
    booster_rate = mean(booster_rate, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    # 分段到 9 个 bin，用于映射深浅色
    rate_bin = cut(booster_rate, breaks = 9, labels = FALSE, include.lowest = TRUE),
    # 获取该 region 的调色板
    base_palette = map(region, ~ region_palettes[[.x]]),
    # 应用深浅色
    fill_color = map2_chr(base_palette, rate_bin, ~ .x[.y])
  )

ggplot(booster_plot_data, aes(x = month, y = state)) +
  geom_tile(aes(fill = fill_color), color = "white") +
  scale_fill_identity() +
  labs(
    title = "Booster Rate Heatmap by State and Month",
    subtitle = "Region hue, Rate darkness",
    x = "Month", y = "State (sorted by avg. Booster Rate)"
  ) +
  theme_minimal(base_size = 12)
```

### LME Models

#### Prepare variables
```{r}
dat = dat |> 
  mutate(
    death_rate = deaths / pop * 1e5,
    case_rate = cases / pop * 1e5,
    hosp_rate = hosp / pop * 1e5,
    vax_rate = series / pop * 1e2,  # percent fully vaccinated
    booster_rate = booster / pop * 1e2  # percent boosted
  )

dat = dat |> 
  group_by(state) |> 
  arrange(date) |> 
  mutate(
    case_rate_lag2 = lag(case_rate, 2),
    hosp_rate_lag2 = lag(hosp_rate, 2)
  ) |> 
  ungroup()

dat_nofirst2 = dat |>
  filter(!is.na(case_rate_lag2), !is.na(hosp_rate_lag2))
```


##### Model 1 lmer
```{r}
model_1 = lmer(death_rate ~ case_rate + hosp_rate + vax_rate + booster_rate + ns(date_num, df = 5) + (1 | state), data = dat_nofirst2)

summary(model_1)


plot(ggpredict(model_1, terms="date_num [all]"))
AIC(model_1)
```



##### Model 2 lag cases hosp
```{r}
model_lag = lmer(death_rate ~ case_rate_lag2 + hosp_rate_lag2 + vax_rate + booster_rate + ns(date_num, df = 8) + (1 | state), data = dat_nofirst2)

summary(model_lag)

plot(ggpredict(model_lag, terms="date_num [all]"))
AIC(model_lag)
```

```{r}
aic_results = tibble(df = 2:10, AIC = NA_real_)

# Loop over degrees of freedom
for (i in 2:10) {
  model = lmer(death_rate ~ case_rate_lag2 + hosp_rate_lag2 + vax_rate + booster_rate +
                 ns(date_num, df = i) + (1 | state), data = dat_nofirst2)
  aic_results$AIC[aic_results$df == i] = AIC(model)
}

print(aic_results)
which.min(aic_results$AIC)
```



```{r}
fit = model_lag
state_to_plot = "CA"

# Remove rows with missing lagged predictors before prediction
dat_pred = dat |>
  filter(!is.na(case_rate_lag2), !is.na(hosp_rate_lag2)) |>
  mutate(predicted = predict(fit))

# Join predictions back to full data (optional, only if you want to keep NA rows in plotting)
dat_state = dat |> 
  filter(state == state_to_plot) |>
  left_join(dat_pred |> select(state, date, predicted), by = c("state", "date"))

# Plot observed vs predicted
ggplot(dat_state, aes(x = date)) +
  geom_point(aes(y = death_rate), color = "red", alpha = 0.6, size = 1.2) +  # Observed
  geom_line(aes(y = predicted), color = "blue", linewidth = 1) +  # Predicted
  labs(title = paste("Observed vs. Predicted Death Rate in", state_to_plot),
       y = "Death Rate (per 100,000)",
       x = "Date") +
  theme_minimal()
```

### Model Diagnostics
```{r}
plot(model_lag)  # base R method for lmer

resid_df = data.frame(
  fitted = fitted(model_lag),
  residuals = resid(model_lag)
)

ggplot(resid_df, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  theme_minimal() +
  labs(title = "Residuals vs. Fitted", x = "Fitted", y = "Residuals")
```

```{r}
qqnorm(resid(model_lag))
qqline(resid(model_lag), col = "blue")

data.frame(resid = resid(model_lag)) |> 
  ggplot(aes(sample = resid)) +
  stat_qq() +
  stat_qq_line(color = "blue") +
  theme_minimal() +
  labs(title = "Q-Q Plot of Residuals")
```
data:image/png;base64,iVBORw0KGgoAAAANSUhEUgAAAAIAAAAUCAYAAACnOeyiAAAAXklEQVR4XmNgAIL///8zMYSGhjIDGYIMIiIMvECGMwMDN4M4kFEDUqIIZKwDMdSBjAsghj6Q8QPEMAAy/lOBoQekv4AYKkDGfgZeXl4RICOLQUtLiw3IUAJJMQIZ7AC2tU2tJWCy/wAAAABJRU5ErkJggg==

### Excess death rate
```{r}
# Predict death rate using the lag model
dat_excess = dat_nofirst2 |> 
  mutate(predicted_death_rate = predict(model_lag),
         excess_death_rate = death_rate - predicted_death_rate)

# View a few rows
head(dat_excess |> select(state, date, death_rate, predicted_death
```_rate, excess_death_rate))

```{r}
state_to_plot = "CA"

dat_excess |> 
  filter(state == state_to_plot) |> 
  ggplot(aes(x = date, y = excess_death_rate)) +
  geom_line(color = "darkred") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_minimal() +
  labs(title = paste("Excess Death Rate Over Time in", state_to_plot),
       y = "Excess Death Rate (per 100,000)",
       x = "Date")
```


